\RequirePackage[l2tabu, orthodox]{nag}
% I hope this turns out well. 
\documentclass[10pt,letter]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[english]{isodate}
\usepackage[parfill]{parskip}
\usepackage[colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{enumerate}
\usepackage{mathtools}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{cancel}
\usepackage{color,soul}
\usepackage[margin=2cm,letterpaper]{geometry}
\usepackage{graphicx}
\usepackage{floatrow}
\usepackage{calc}
\usepackage{xfrac}
\usepackage[font=small]{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\SetTracking{encoding={*}, shape=sc}{40}
\usepackage[tocindentauto]{tocstyle}


\microtypecontext{spacing=nonfrench}

\title{GRE Physics Study Notes - Strategy / Tricky Notes}
\author{Courtesy of Nicole Duncan, transcribed by Jeren Suzuki}
\date{Last Edited \today}

\begin{document}

\maketitle
\pagenumbering{Roman}
\newpage
\tableofcontents
\newpage
\pagenumbering{arabic}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Look at question before trying to answer the question}
\begin{enumerate}
    \item How far apart are they? \\ 
    Orders of magnitude, drop all but 1st significant fig\\
    very close $\rightarrow$ need precise values
    \item ``Most nearly'' means that my exact answer will not appear\\
    Can do some rounding, keep 1 or 2 significant figs
    \item True vs Not True: \\
    Make sure you know which one\\
    Circle the words
\end{enumerate}

\section{Don't just guess} % (fold)
\label{sec:don_t_just_guess}
Try to write an equation or equality before answering. Justify short answers.
% section don_t_just_guess (end)

\section{Use a Section Classification Key} % (fold)
\label{sec:use_a_section_classification_key}
\begin{enumerate}
    \item Circle \# $\rightarrow$ come back definitely 
    \item Cross $\rightarrow$ done
\end{enumerate}
% section use_a_section_classification_key (end)

\section{Read the Whole Problem} % (fold)
\label{sec:read_the_whole_problem}
Make sure I understand problem very quickly.
% section read_the_whole_problem (end)

\section{KNOW WHICH ONES TO SKIP} % (fold)
\label{sec:know_which_ones_to_skip}
% section know_which_ones_to_skip (end)

\section{Be Aware Of} % (fold)
\label{sec:units_to_be_aware_of}
\subsection{Units} % (fold)
\label{sub:units}
kcal = $10^3$ cal\\
Answers + given in problem\\
Will ask for answer in kJ, etc.
% subsection units (end)

\subsection{Frequency vs $\omega$} % (fold)
\label{sub:freq}
Hz always means frequency\\
rad/s = $\omega$\\
$\omega = 2 \pi \nu$, units of $\frac{2\pi \textrm{radians} }{\textrm{rev}}$\\
Revolutions always imply frequency
% subsection freq_ (end)

\subsection{Signs and Directions} % (fold)
\label{sub:signs_and_directions}
Especially in conservation of momentum problems, $v$ vs $-v$\\
Doppler Shift Problems: $-v$ if moving towards you, $+v$ if you move towards it
% subsection signs_and_directions (end)


Sometimes they don't specify where $\hat{z}$ is, so don't get hung up on RHR if they don't label it.\\
\subsection{Answer doesn't fit answers} % (fold)
\label{sub:answer_doesn_t_fit_answers}
It might be given in another form. e.g.,
\begin{equation}
    \frac{e^{-\epsilon/kT}}{1 + e^{-\epsilon/kt}} \rightarrow \frac{1}{e^{\epsilon/kT} + 1}
\end{equation}
% subsection answer_doesn_t_fit_answers (end)\\

\section{Problem Strategy} % (fold)
\label{sec:problem_strategy}

\subsection{When in doubt about E\&M fields} % (fold)
\label{sub:subsection_name}
Write out maxwell's equations and look for how the field will be changed or how it'll look. \\
Write out Lorentz forces, how does the E \& B move the particle 
% subsection subsection_name (end)

\subsection{Graphs} % (fold)
\label{sub:graphs}
Pick out key features\\
Axis + relations to other quantities
% subsection graphs (end)

\subsection{Square Roots and Fractions} % (fold)
\label{sub:square_roots_and_fractions}
Use bounding values\\
$x = \sqrt{.78}~,~ .8^2 = .64~,~ .9^2 = .81$, so between 8 and 9.
% subsection square_roots_and_fractions (end)

\subsection{Sneaky Famous Experiments} % (fold)
\label{sub:snenaky_famous_experiments}
They hide them in other questions about similar equations
% subsection snenaky_famous_experiments (end)

\subsection{Numbers go in Last} % (fold)
\label{sub:numbers_go_in_last}
Clearly write exponentials\\
Say \# over to myself as I write
% subsection numbers_go_in_last (end)

\subsection{If the answer is an Equation} % (fold)
\label{sub:if_the_answer_is_an_equation}
Think in limits and units
% subsection if_the_answer_is_an_equation (end)

% section problem_strategy (end)


\section{Common Errors} % (fold)
\label{sec:common_errors}
\subsection{Angstrom} % (fold)
\label{sub:angstrom}
\AA $= 1 \times 10^{-10}$ m
% subsection angstrom (end)

\subsection{Collision Problems} % (fold)
\label{sub:collision_problems}
Conserve $p$ for initial/final conditions - don't forget $\pm$v. \\
Conserve $KE$ for $\Delta E$ between $U$ \& $KE$ only!
% subsection collision_problems (end)

\subsection{Most Nearly} % (fold)
\label{sub:most_nearly}
My answer will not be there, don't fret.
% subsection most_nearly (end)

\subsection{When Rounding} % (fold)
\label{sub:when_rounding}
Keep track of up vs down so will know if final answer is less or more.
% subsection when_rounding (end)

\subsection{Fractions!} % (fold)
\label{sub:fractions_}
\begin{equation}
    \frac{1}{2 \times 10^{-5}} = \frac{1}{2}\times 10^{5} = .5 \times 10^{5} = 5 \times 10^4
\end{equation}
% subsection fractions_ (end)

% section common_errors (end)

\section{Tricks} % (fold)
\label{sec:tricks}
For relativity, write down equations first, then see if you can simply divide. Given $E$ and $p$, find $v$:

\begin{equation}
\frac{E}{p} = \frac{\gamma mc^2}{\gamma mv} \rightarrow v
\end{equation}
% section tricks (end)

\end{document}